---
title: "Lab 3"
format: html
editor: visual
---

## Lab 3

### 1.1 - Loading in data

```{r}
# packages 
library(readr)
library(tidyverse)
library(astsa)
library(forecast)

# reading in data 
wine.csv <- read.table("~/Desktop/P174/monthly-australian-wine-sales-th.csv",
                                               sep = ',',
                                               header = F,
                                               skip = 1,
                                               nrows = 174)
colnames(wine.csv) <- c("Month","Sales")

# making tine series data 
wine <- ts(wine.csv$Sales, start = c(1980,1), frequency = 12)

# plotting time series
ts.plot(wine, main = 'Wine Sales over Time')
```

### 1.2 - Questions

```{r}
# 1 - sample differencing
wine_diff <- diff(wine, lag = 1, differences = 1)

ts.plot(wine_diff, main = 'Difference of Wine') %>%
  abline(h = 0)
```

```{r}
# 2 - acf and pacf of differenced series
wine_diff_acf <- acf(wine_diff, lag.max = 60, plot = T)

wine_diff_pacf <- pacf(wine_diff, lag.max = 60, plot = T)
```

The acf and pacf for the differenced series tell us there is still some seasonality in the data because of the repeating values within the plots. This patterns tells us that the seasonality of the data is probably longer than one month, so the differenced series should have more lag between observations to remove the seasonality.

```{r}
# 3 - 12 month difference
wine_diff <- diff(wine, lag = 12, differences = 1)

# plotting new differenced series
ts.plot(wine_diff, main = 'Difference of Wine') %>%
  abline(h = 0)

# acf and pacf
wine_diff_acf <- acf(wine_diff, lag.max = 60, plot = T)
wine_diff_pacf <- pacf(wine_diff, lag.max = 60, plot = T)
```

### Modeling GNP

```{r}
# 1 - plotting gnp as time series 
ts.plot(gnp, main = "US GNP")

# converting gnp into csv for python
gnp_data <- gnp
gnp_df <- data.frame(
  Year = time(gnp_data),
  GNP = as.numeric(gnp_data)
)

write.csv(gnp_df, "gnp_data.csv", row.names = FALSE)
```

The data shows an exponentially-increasing trend over time which implies the data isn't stationary. To make the data stationary, we could try a log transformation of the differences in quarters.

```{r}
gnp_transform <- log(gnp) %>%
  diff(lag = 1, differences = 1)

ts.plot(gnp_transform, main = "US GNP")
```

The time series now appears stationary.

```{r}
# 3 - plotting acf and pacf for new series
acf2(gnp_transform, plot = T)
```

There appears to be a pattern in the acf values for the time series with a period of h = 8 (since lag on x-axis of the plots is divided by 4). There doesn't appear to be a pattern for the pacf values.

```{r}
# 4 - making MA(2) model and finding coefficients
ma.model <- Arima(gnp_transform, order = c(0,0,2))

summary(ma.model)

ma.model$coef
```

The coefficients for the model are (approximately) $\theta_1$ = 0.303 and $\theta_2$ = 0.204

```{r}
# 5 - plotting residuals and acf of MA(2)
plot(ma.model$residuals)

acf(ma.model$arma, plot = T)
```

The model does appear to be a good fit because the residuals are decreasing as time passes.

```{r}
# 6 - 
```
